---
title: "Visualising the trajectory"
output:
  blogdown::html_page:
    toc: true
weight: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, results="hold")
```

```{r, results='hide', message=FALSE, warning=FALSE}
library(dyno)
library(tidyverse)
```

The main functions for plotting a trajectory are included in the [dynplot package](/reference/dynplot).

We'll use an example toy dataset
```{r}
set.seed(1)
dataset <- dyntoy::generate_dataset(model = "bifurcating", num_cells = 200)
```

To visualise a trajectory, you have to take into account two things:

* Where will I place the trajectory and cells in my 2D space
* What do I want to visualise along the trajectory based on color

Depending on the answer on these two questions, you will need different visualisations:

```{r cells, fig.width=15, fig.height=10, echo=FALSE, message=FALSE}
feature_oi <- dataset$feature_info$feature_id[[1]]
grouping <- group_onto_nearest_milestones(dataset)

empty <- ggplot() + theme_void()
question <- empty + geom_text(aes(0, 0, label="????"))
unavailable <- empty + geom_text(aes(0, 0, label="----"))
titles <- ggplot() + geom_text(aes(x=1, y = 0.5, label = "Color"), hjust = 1) + geom_text(aes(x=0.5, y = 0, label = "Position")) + geom_segment(aes(x=0.2, y=0.8, xend=1, yend=0), color = "#666666") + theme_void() + scale_x_continuous(limits = c(0, 1)) + scale_y_continuous(limits = c(0, 1))
title <- list(
  titles, 
  # empty + geom_label(aes(0, 0, label="Topology")),
  empty + geom_label(aes(0, 0, label="Ordering")),
  empty + geom_label(aes(0, 0, label="Cell grouping")),
  empty + geom_label(aes(0, 0, label="Feature expression")), 
  empty + geom_label(aes(0, 0, label="Pseudotime"))
)
dendro <- list(
  empty + geom_label(aes(0, 0, label="Dendrogram")),
  # plot_dendro(dataset),
  plot_dendro(dataset, "milestone"),
  plot_dendro(dataset, grouping=grouping, groups=groups),
  plot_dendro(dataset, feature_oi=feature_oi),
  plot_dendro(dataset, "pseudotime")
)
onedim <- list(
    empty + geom_label(aes(0, 0, label="Onedim")),
    # plot_onedim(dataset),
    plot_onedim(dataset, "milestone"),
    plot_onedim(dataset, grouping=grouping, groups=groups),
    plot_onedim(dataset, feature_oi=feature_oi),
    plot_onedim(dataset, "pseudotime")
  )
graph <- list(
  empty + geom_label(aes(0, 0, label="Graph")),
  # plot_graph(dataset),
  plot_graph(dataset, "milestone"),
  plot_graph(dataset, grouping=grouping, groups=groups),
  plot_graph(dataset, feature_oi=feature_oi),
  plot_graph(dataset, "pseudotime")
)
dimred <- list(
  empty + geom_label(aes(0, 0, label="Dimensionality reduction")),
  # plot_dimred(dataset),
  plot_dimred(dataset, "milestone"),
  plot_dimred(dataset, grouping=grouping, groups=groups),
  plot_dimred(dataset, feature_oi=feature_oi),
  plot_dimred(dataset, "pseudotime")
)
patchwork::wrap_plots(
  c(title, dendro, onedim, graph, dimred),
  nrow=5,
  byrow=TRUE
) & theme(legend.position = "none")
```

## Visualising the trajectory on a dimensionality reduction

The most common way to visualise a trajectory is to plot it on a dimensionality reduction of the cells. Often (but not always), a TI method will already output a dimensionality reduction itself, which was used to construct the trajectory. For example:

```{r}
model <- infer_trajectory(dataset, ti_mst())
head(get_dimred(model), 5)
```
Dynplot will use this dimensionality reduction if available, otherwise it will calculate a dimensionality reduction under the hood:

```{r}
plot_dimred(model)
```

You can also supply it with your own dimensionality reduction. In this case, the trajectory will be _projected_ onto this dimensionality reduction.
```{r}
dimred <- dyndimred::dimred_umap(dataset$expression)
plot_dimred(model, dimred = dimred)
```

On this plot, you can color the cells according to 

1. **Cell ordering**. In which every milestone gets a color and the color changes gradually between the milestones. This is the default.
2. **Cell grouping**. A character vector mapping a cell to a group. In this example, we group the cells to their nearest milestone, but typically you will want to supply some external clustering here.
3. **Feature expression**. In which case you need to supply the `expression_source` which is usually the original dataset, but can be in any format accepted by `dynwrap::get_expression()`.
4. **Pseudotime**. The distance to a particular root milestone. The [adapting guide](../5-adapting) discusses how to change the root of a trajectory.

```{r, fig.height = 10, fig.width = 10}
patchwork::wrap_plots(
  plot_dimred(model) + ggtitle("Cell ordering"),
  plot_dimred(model, grouping = group_onto_nearest_milestones(model)) + ggtitle("Cell grouping"),
  plot_dimred(model, feature_oi = "G1", expression_source = dataset) + ggtitle("Feature expression"),
  plot_dimred(model, "pseudotime", pseudotime = calculate_pseudotime(model)) + ggtitle("Pseudotime")
)
```

<div class="notices tip">

The result is just a regular ggplot2 object, so you can change and adapt it further:

```{r}
plot_dimred(model) + 
  ggtitle("Cells are amazing") + 
  annotate("label", x = model$dimred[1, 1], y = model$dimred[1, 2], label = "Amazing cells") +
  annotate("label", x = model$dimred[2, 1], y = model$dimred[2, 2], label = "Also amazing cells")
```


</div>

<div class="notices future">
Plotting of RNA velocity on top of a trajectory would be very _very_ useful, but is not yet included.
See the issue on github: https://github.com/dynverse/dynplot/issues/29

</div>

## Plotting the trajectory itself

If the dataset is too complex to be visualised using a 2D dimensionality reduction, it can be useful to visualise the trajectory itself. We provide 3 ways to do this:

### Plotting in a dendrogram

When the trajectory has a tree structure and a clear direction, it is often the most intuitive to visualise it as a dendrogram:

```{r}
plot_dendro(model)
```

This visualisation hinges on the correct location of a root, which will be further discussed in the [adapting guide](../5-adapting).

Here you can again color the cells in different ways similar as `plot_dimred`, i.e.:

```{r}
plot_dendro(model, "pseudotime")
```

### Plotting as a graph

For more complex topologies, which include cycles or disconnected pieces, the trajectory can be visualised as a 2D graph structure.

```{r}
disconnected_dataset <- dyntoy::generate_trajectory(model = "disconnected", num_cells = 300)
plot_graph(disconnected_dataset)
```

### Plotting in one dimension

Sometimes it can be useful to visualise the trajectory in one dimension, so that you can use the other dimension for something else:

```{r}
plot_onedim(add_root(model))
```


## Visualising many genes along a trajectory

<div class="notices warning">
Heatmap plotting is currently still experimental.
</div>

A one-dimensional visualisation is especially useful if you combine it with a heatmap:

```{r}
plot_heatmap(model, expression_source = dataset)
```

Selecting relevant features for this heatmap is discussed in a [later guide](../6-tde), but suffice it to say that `plot_heatmap()` by default will plot those features that best explain the main differences over the whole trajectory.
