---
title: "Adapting the trajectory"
output:
  blogdown::html_page:
    toc: true
weight: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, results="hold")
```

```{r, results='hide', message=FALSE, warning=FALSE}
library(dyno)
library(tidyverse)
```

## Simplifying

Intermediate milestones can be removed by simplyfing the trajectory:

```{r}
model <- dyntoy::generate_dataset(model = dyntoy::model_linear(num_milestones = 10))
simplified <- simplify_trajectory(model)
```

```{r, echo = FALSE, message = FALSE, fig.height = 3, fig.width = 10}
patchwork::wrap_plots(
  plot_onedim(model, label_milestones = TRUE) + ggtitle("Original"),
  plot_onedim(simplified, label_milestones = TRUE) + ggtitle("Simplified")
)
```



## Rooting

TI methods often do not have an idea where the root of a trajectory is. We provide two ways of rooting a trajectory. After rooting, all other edges will point away from the root.

```{r}
set.seed(1)
model <- dyntoy::generate_dataset(model = dyntoy::model_bifurcating())
```

### Manually

If you know the milestone (or cell) that is at the start of the trajectory, you can directly call `add_root`:

```{r}
model_rooted <- model %>% add_root(root_milestone_id = "M4")
```

```{r, echo = FALSE, message = FALSE, fig.width = 10, fig.height = 5}
patchwork::wrap_plots(
  plot_dimred(model, label_milestones = TRUE) + ggtitle("Original"),
  plot_dimred(model_rooted, label_milestones = TRUE) + ggtitle("Rooted at M4")
)
```


### Using marker genes

If you know some marker genes that are highly expressed at the start of the trajectory, rooting can be done implicitely:

```{r}
model_rooted <- model %>% add_root_using_expression("G1", expression_source = model)
```

```{r, echo = FALSE, message = FALSE, fig.width = 10, fig.height = 5}
patchwork::wrap_plots(
  plot_dimred(model, label_milestones = TRUE, feature_oi = "G1") + ggtitle("Original"),
  plot_dimred(model_rooted, label_milestones = TRUE, feature_oi = "G1") + ggtitle("Rooted using G1 expression")
)
```

<div class="notices future">
Rooting a trajectory based on RNA velocity is on our todo list. See https://github.com/dynverse/dynwrap/issues/115
</div>

## Annotating

<div class="notices warning">
Annotating/labelling milestones is still experimental
</div>

Similarly as with rooting, there are also two ways to annotate the milestones within a trajectory:

### Manually

```{r}
model_labelled <- model %>% label_milestones(c(M1 = "start", M2 = "end1", M3 = "decision", M4 = "end2"))
```

```{r, echo = FALSE, message = FALSE, fig.width = 5, fig.height = 5}
plot_dimred(model_labelled, label_milestones = TRUE) + ggtitle("Labelled")
```


### Using marker genes

```{r}
model_labelled <- label_milestones_markers(
  model,
  markers = list(
    G1high = c("G1"),
    G5high = c("G5"),
    G3high = c("G3"),
    G8high = c("G8")
  )
)
```

```{r, echo = FALSE, message = FALSE, fig.width = 5, fig.height = 5}
plot_dimred(model_labelled, label_milestones = TRUE, feature_oi = "G1") + ggtitle("Labelled")
```

<div class="notices future">
Annotating milestones based on external information is on our todo list.
</div>

## Limitations

<div class="notices future">
Splitting a trajectory and adding intermediate milestones is on our todo list.
</div>
